Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-05 paper.tex
more

Clearly, this definition of list membership tells us not just that an item is in
a list, but also its \emph{position} in that list.
That piece of information gives use decidable equality, restricting the
predicate to discrete sets.
In section~\ref{manifest-enumerability}, we will show how to hide the position,
giving us a variant of the predicate which allows for non-set finite types.
\subsection{Proving Equivalence with Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
Here, we will formalise this equivalence, by first proving an equivalence with a
container-based definition of split enumerability.

Containers \cite{abbottContainersConstructingStrictly2005} give us a general way
to construct functors from generic building-blocks.
\begin{definition}[Container]
  A container is a dependent pair of shapes, and positions indexed by those
  shapes.
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
\end{definition}

For instance, the shape of lists is given by \(\mathbb{N}\), and the positions are then
\(\mathbf{Fin}\)s indexed by that \(\mathbb{N}\).
If we swap out wherever we used ``list'' in the definition of split
enumerability with this container-based list, we get an equivalent finiteness
predicate.
\begin{lemma} \label{container-split-enum-equiv}
  The container-based definition of split enumerability is equivalent to the
  inductive type version.
  \begin{equation}
    \sum_{(\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket A)} , \prod_{(x : A)} , x \in xs \simeq
    \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
  Membership for container-defined lists is a fiber.
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{lemma}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}

This version of split enumerability is equivalent to a split surjection from a
finite prefix of the natural numbers. 
\begin{lemma} \label{container-split-enum-surj}
  Container-defined split enumerability is equivalent to a split surjection from
  some \(\mathbf{Fin}\).
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \mathbf{Fin}(n) \twoheadrightarrow ! A
  \end{equation}
\end{lemma}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
This lets us prove our overall goal for this section.
\begin{theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
\end{theorem}
\begin{proof}
  This follows from lemmas \ref{container-split-enum-equiv} and
  \ref{container-split-enum-surj}.
\end{proof}
\subsection{Decidable Equality}
\begin{lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{lemma}
\begin{proof}
  We use a corollary, proven in the appendix, that if there is a
  split-surjection from \(A\) to \(B\), and \(A\) is discrete, then \(B\) is
  also discrete.
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{discrete}
  \todo[inline]{Rework this proof to rely on the split surjection equivalence
    from before}
\end{proof}

This also tells us that any split enumerable type must be a set.
\begin{lemma}
  Any split enumerable type is a set.
\end{lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{lemma}
  \(\bot\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{lemma}
  \(\top\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{theorem}
  Split-enumerability is closed under \(\sum\).
\end{theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
While manifest Bishop finiteness may seem stricter than split enumerability, we
will see soon that it is actually equivalent in strength.
Why study it separately to split enumerability at all, then?
What purpose does it serve on our road to Kuratowski finiteness?
Those questions are answered by the following theorem:
\begin{lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  \todo[inline]{Finish prose for this proof}
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{lemma}
  Any manifest Bishop finite set is split enumerable.
\end{lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-to-split-enum}
\end{proof}
\begin{theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

  It suffices now to prove the following:
  \begin{equation}
    \prod_{(x : A)} , \prod_{(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
  And from that we can generate our cover proof.

  Full details of the formal proof are provided in the appendix.
\end{proof}
Note that while manifest Bishop finiteness as split enumerability are equivalent
in ``strength'', the two types are not equivalent.
In particular, there are infinitely many inhabitants of \(\mathcal{E}!\), while
for a type \(A\) with \(n\) inhabitants, there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{theorem}
  Manifest bishop finiteness is closed over dependent functions (\(\prod\)-types).

>>>>

Clearly, this definition of list membership tells us not just that an item is in
a list, but also its \emph{position} in that list.
That piece of information gives use decidable equality, restricting the
predicate to discrete sets.
In section~\ref{manifest-enumerability}, we will show how to hide the position,
giving us a variant of the predicate which allows for non-set finite types.
\subsection{Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
In this section, we will prove that equivalence, formally.

\begin{theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{theorem}
\begin{proof}
  This proof is the composition of two intermediate equivalences: lemmas
  \ref{container-split-enum-equiv} and \ref{container-split-enum-surj}.
  This first of these proves that our definition of finiteness is equivalent to
  another definition which uses lists defined as a \emph{container}, rather than
  the usual inductive type definition.
  The second of these shows that the container-based finiteness predicate is
  equivalent to a split surjection from some \(\mathbf{Fin}\).
\end{proof}

To do so, we will first need to take a small detour to look at Containers.
Containers \cite{abbottContainersConstructingStrictly2005} give us a general way
to construct functors from generic building-blocks.
\begin{definition}[Container]
  A container is a dependent pair of shapes, and positions indexed by those
  shapes.
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{definition}

For instance, the shape of lists is given by \(\mathbb{N}\), and the positions are then
\(\mathbf{Fin}\)s indexed by that \(\mathbb{N}\).
If we swap out wherever we used ``list'' in the definition of split
enumerability with this container-based list, we get an equivalent finiteness
predicate.
\begin{lemma} \label{container-split-enum-equiv}
  The container-based definition of split enumerability is equivalent to the
  inductive type version.
  \begin{equation}
    \sum_{(\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket A)} , \prod_{(x : A)} , x \in xs \simeq
    \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
\end{lemma}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}

This version of split enumerability is equivalent to a split surjection from a
finite prefix of the natural numbers. 
\begin{lemma} \label{container-split-enum-surj}
  Container-defined split enumerability is equivalent to a split surjection from
  some \(\mathbf{Fin}\).
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{discrete}
\end{proof}
This also tells us that any split enumerable type must be a set.
\begin{lemma}
  Any split enumerable type is a set.
\end{lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{lemma}
  \(\bot\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{lemma}
  \(\top\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{theorem}
  Split-enumerability is closed under \(\sum\).
\end{theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
While manifest Bishop finiteness may seem stricter than split enumerability, we
will see soon that it is actually equivalent in strength.
Why study it separately to split enumerability at all, then?
What purpose does it serve on our road to Kuratowski finiteness?
Those questions are answered by the following theorem:
\begin{lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, we can define a container-based version of bishop finiteness
  which is a reassociation of a contractible map, and then show that the
  container-based predicate is equivalent to the list-based.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{lemma}
  Any manifest Bishop finite set is split enumerable.
\end{lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-to-split-enum}
\end{proof}
\begin{theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

  It suffices now to prove the following:
  \begin{equation}
    \prod_{(x : A)} , \prod_{(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
  And from that we can generate our cover proof.
\end{proof}
Note that while manifest Bishop finiteness as split enumerability are equivalent
in ``strength'', the two types are not equivalent.
In particular, there are infinitely many inhabitants of \(\mathcal{E}!\), while
for a type \(A\) with \(n\) inhabitants, there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{theorem}
  Manifest bishop finiteness is closed over dependent functions (\(\prod\)-types).
